Incremental Faint Object Detection in Multi-Epoch Observations

### 2016-11-02 ms.tex
Sxn 4 fixes; Sxn 5 line breaks
Fix punctuation in App

\begin{figure}
\epsscale{1.2}
\plotone{fig/f8u-U.png}
\caption{The Bayes factor can differentiate significantly better between real objects and noise peaks when we include astrometric information. Here we plot logarithm of product of the photometric and astrometric Bayes factors, the latter which comes directly from the cross-identification based on the celestial coordinates, see text for details.}
\label{fig:bf2}
\end{figure}

@@

We will expand the field and its gradient to second order around a peak at the 
position $\rr_p$.
\begin{eqnarray}
	f(\rr) = f(\rr_p)+\frac{1}{2} g_{ij}(\rr_p) (\rr-\rr_p)_i  (\rr-\rr_p)_j\\
	h_i(\rr) = (\rr-\rr_p)_j g_{ij}(\rr_p)
\end{eqnarray}
where we already use the fact that the gradient of the field at a peak is zero, 
i.e. $h_i(\rr_p)=0$. Provided that $g$ is non-singular at $\rr_p$, we can express 
$\rr-\rr_p$ from the second equation:
\begin{equation}
	\rr-\rr_p = g^{-1}(\rr_p) \hh(\rr_p).
\end{equation}
We can write a Dirac delta that picks all extremal points of $f$ as
\begin{equation}
	\delta^{(2)}(\rr-\rr_p) = |\det g(\rr)| \delta^{(2)}[\hh(\rr)].
\end{equation}
This expression turns a continous random field, defined at all points over our 
two-dimensional space into a discrete point process, that of the extremal points 
of the field,
\begin{equation}
	n_{\rm{}ext}(\rr) = |\det g(\rr)| \delta^{(2)}[\hh(\rr)].
\end{equation}

@@

We can now explicitely write down the correlation matrix $C$ of 
${\bf v} = (f,g_{11},g_{12},g_{22})$ and $H$ for $\hh= (h_1, h2)$, as
\begin{equation}
	 C = \left(
		\begin{array}{cccc}
			\sigma_0^2 & -\sigma_1^2/2 & 0 & -\sigma_1^2/2\\
			-\sigma_1^2/2 & 3\sigma_2^2/8 & 0 & \sigma_2^2/8\\
			  0 & 0 & \sigma_2^2/8  & 0\\
			-\sigma_1^2/2 & \sigma_2^2/8 & 0 & 3\sigma_2^2/8
		 \end{array}\right),
\end{equation}
\begin{equation}
	 H = \left(
		\begin{array}{cccc}
			-\sigma_1^2/2 & 0 \\
			0 & -\sigma_1^2/2
		 \end{array}\right).
\end{equation}
With these we can write the multivariate Gaussian distribution using the inverse 
of the correlation matrix as a product of two independent distributions
\begin{equation}
	dP =\exp\left(-\frac{\vv^T C^{-1}\vv}{2}\right)
  			\frac{d^4 \vv} {(2\pi)^2|C|^{1/2}}
		 \exp\left(-\frac{\hh^T H^{-1} \hh}{2}\right) 
			\frac{d^2 \hh} {2\pi|H|^{1/2}}
\end{equation}
Before we proceed further, the second derivative tensor can more conveniently
described with the two eigenvalues $\lambda_1,\lambda_2$ and a rotation angle 
$\phi$, as
\begin{eqnarray}
	g_{11} =& \lambda_1 \cos^2\phi +\lambda_2 \sin^2\phi\\
	g_{12} =& (\lambda_1-\lambda_2) \sin\phi \cos\phi\\
	g_{22} =& \lambda_1 \sin^2\phi + \lambda_2 \cos^2\phi
\end{eqnarray}
For simplicity let us introduce the dimensionless variables 
$x=(\lambda_1+\lambda_2)/\sigma_2$, the trace of the second derivative tensor, 
$y=(\lambda_1-\lambda_2)/\sigma_2$, and $z = f/\sigma_0$. The Jacobian of the 
transformation from $(f,g_{11},g_{12},g_{22})$ to $(z,x,y,\phi)$ is
\begin{equation}
		J = \sigma_0\sigma_2^3 y/2.  
\end{equation} 
Let us also introduce the dimensionless $\gamma$ and the characteristic scale 
$R$ as
\begin{equation}
	\gamma = \frac{\sigma_1^2}{\sigma_0\sigma_2},
		\qquad R^2 = \frac{\sigma_1^2}{\sigma_2^2}.
\end{equation}
The quadratic form containing $\vv$ in the exponent can be written with the new 
variables as
\begin{equation}
	Q = v^T C^{-1} v = \left(
		2 y^2	+\frac{x^2 + 2\gamma x z + z^2}{1-\gamma^2}\right)
\end{equation}
The determinants of $C$ and $H$ are 
\begin{equation}
	|C| = \frac{1}{64}(1-\gamma^2)\sigma_0^2\sigma_2^6,\qquad
	|H| = \frac{1}{4} \sigma_1^2
\end{equation}
In these variables, the unconstrained probability distribution for 
$(x,y,z,\phi)$ becomes
\begin{equation}
	dP = \frac{4y}{(2\pi)^2 \sqrt{1-\gamma^2}} \exp(-\frac{1}{2}Q) 
		\,dx\ dy\ dz\ d\phi
\end{equation}
In order to properly handle the symmetries of the problem, we can assume that 
$\lambda_1\geq\lambda_2$. Then still any $(\lambda_1,\lambda_2)$ pair can be 
mapped onto itself by a 180 degree rotation, so the valid range of $\phi$ is 
$(0,\pi)$.  Since none of the terms depend on $\phi$, we can integrate 
over $\phi$, resulting in
\begin{equation}
	dP =  
		 \exp\left[-\frac{x^2 + 2\gamma x z + z^2}{2(1-\gamma^2)}\right] 
			\frac{\,dx\ dz}{2\pi \sqrt{1-\gamma^2}}
		\left( e^{-y^2}\,2y\ dy\right)
\end{equation}
The constraint $\lambda_1>\lambda_2$ maps onto $0<y$. If we perform the 
integration over $y>0$, and $x,z$ over $(-\infty,\infty)$, we get 1, as we 
should, for the unconstrained probability for a general point.

As we introduce the peak constraints, we need to first consider the impact
on the gradient. The constrained probability distribution
is
\begin{equation}
	dw_x = \exp\left(-\frac{\hh^T H^{-1} \hh}{2}\right) 
			|\det g| \delta^{(2)}[\hh]			
			\frac{d^2 \hh} {2\pi|H|^{1/2}}
\end{equation}
After integrating over $d^2\hh$ we get the extremum weight
\begin{equation}
	w_x = \frac{|\det g|}{2\pi|H|^{1/2}} =\frac{x^2-y^2}{4 \pi R^2}
\end{equation}
This will multiply the unconstrained probability for the density of extremal 
points of the random field, 
\begin{equation}
		dn_{\rm{}pk}=\left[\frac{(x^2-y^2)y}{2\pi R^2} e^{-y^2}\,dy\right]
		\exp\left[-\frac{x^2 + 2\gamma x z + z^2}{2(1-\gamma^2)}\right] 
			\frac{\,dx\ dz}{2\pi \sqrt{1-\gamma^2}}
\end{equation}
For a peak both eigenvalues of the second derivative tensor must be negative. 
In the rotated coordinates $(x,y)$, this means that $x<0$, and $0<y<-x$.
We can easily integrate over the allowed range of $y$ next, yielding
\begin{equation}
	\int_0^{-x} dy\, y\/ (x^2-y^2)\, e^{-y^2} 
		=\frac{1}{2}(x^2-1+e^{-x^2}).
\end{equation}
We are left with
\begin{equation}
	dn_{\rm{}pk} = \frac{(-1 + x^2 +e^{-x^2})}{4\pi R^2}  
	\exp\left[-\frac{x^2 + 2 x z \gamma + z^2 }{2 (1-\gamma^2)}\right]
	\frac{\,dx\ dz}{2\pi \sqrt{1-\gamma^2}}
\end{equation}
Let us introduce the function $B(x)$ as
\begin{equation}
	B(s,b) = \sqrt{\frac{\pi}{b}} \exp\left({\frac{s^2}{2b}}\right)
	\left[1+ {\rm erf}\left( \frac{s}{\sqrt{2b}}\right)\right],
\end{equation}
Evaluating the integral over $-\infty<x\leq 0$ in Mathematica, we obtain
\begin{equation}
 n_{\rm{}pk}(s,\gamma)= 
		%\exp\left(-\frac{s^2}{2\gamma^2}\right) 
		\frac{e^{-\frac{s^2}{2\gamma^2}}}{8\pi^2R^2}
		\Bigg[
			(1-\gamma^2)s +[s^2 - \gamma^2 (1 + s^2)] B(s,1)
			+B(s, 3- 2\gamma^2)
		\Bigg]
\end{equation}
with
\begin{equation}
	s = \frac{\gamma z}{\sqrt{1-\gamma^2}}
\end{equation}

We need to evaluate the shape parameter $\gamma$. Assume that the window 
function applied to the random field is a Gaussian with a scale $a$,
\begin{equation}
	w(r) = \frac{1}{2\pi a^2} \exp\left(-\frac{r^2}{2 a^2}\right).
\end{equation}
Its Fourier transform is also a Gaussian,
\begin{equation}
	W(k) = \exp\left(-\frac{k^2a^2}{2}\right).
\end{equation}
We model the sky noise as a white noise with a flat spectrum. Thus the
correlations in the measured random field are determined by the window function,
i.e.,
\begin{equation}
	P(k) = A\,|W(k)|^2 = A \exp\left(-k^2a^2\right)
\end{equation}
With this power spectrum it is trivial to compute the scale and the shape 
parameters as
\begin{equation}
	\gamma^2 = \frac{1}{2}, \qquad R^2 = \frac{a^2}{2}
\end{equation}

>>>>

\begin{figure}[t!]
\epsscale{1.2}
\plotone{fig/f8u-U.png}
\caption{The Bayes factor can differentiate significantly better between real objects and noise peaks when we include astrometric information. Here we plot logarithm of product of the photometric and astrometric Bayes factors, the latter which comes directly from the cross-identification based on the celestial coordinates, see text for details.}
\label{fig:bf2}
\end{figure}

@@

We will expand the field and its gradient to second order around a peak at the 
position $\rr_p$:
\begin{eqnarray}
	f(\rr) = f(\rr_p)+\frac{1}{2} g_{ij}(\rr_p) (\rr-\rr_p)_i  (\rr-\rr_p)_j\\
	h_i(\rr) = (\rr-\rr_p)_j g_{ij}(\rr_p),
\end{eqnarray}
where we already use the fact that the gradient of the field at a peak is zero, 
i.e. $h_i(\rr_p)=0$. Provided that $g$ is non-singular at $\rr_p$, we can express 
$\rr-\rr_p$ from the second equation:
\begin{equation}
	\rr-\rr_p = g^{-1}(\rr_p) \hh(\rr_p).
\end{equation}
We can write a Dirac delta that picks all extremal points of $f$ as
\begin{equation}
	\delta^{(2)}(\rr-\rr_p) = |\det g(\rr)| \delta^{(2)}[\hh(\rr)].
\end{equation}
This expression turns a continous random field, defined at all points over our 
two-dimensional space into a discrete point process, that of the extremal points 
of the field,
\begin{equation}
	n_{\rm{}ext}(\rr) = |\det g(\rr)| \delta^{(2)}[\hh(\rr)].
\end{equation}

@@

We can now explicitely write down the correlation matrix $C$ of 
${\bf v} = (f,g_{11},g_{12},g_{22})$ and $H$ for $\hh= (h_1, h2)$, as
\begin{equation}
	 C = \left(
		\begin{array}{cccc}
			\sigma_0^2 & -\sigma_1^2/2 & 0 & -\sigma_1^2/2\\
			-\sigma_1^2/2 & 3\sigma_2^2/8 & 0 & \sigma_2^2/8\\
			  0 & 0 & \sigma_2^2/8  & 0\\
			-\sigma_1^2/2 & \sigma_2^2/8 & 0 & 3\sigma_2^2/8
		 \end{array}\right),
\end{equation}
\begin{equation}
	 H = \left(
		\begin{array}{cccc}
			-\sigma_1^2/2 & 0 \\
			0 & -\sigma_1^2/2
		 \end{array}\right).
\end{equation}
With these we can write the multivariate Gaussian distribution using the inverse 
of the correlation matrix as a product of two independent distributions
\begin{equation}
	dP =\exp\left(-\frac{\vv^T C^{-1}\vv}{2}\right)
  			\frac{d^4 \vv} {(2\pi)^2|C|^{1/2}}
		 \exp\left(-\frac{\hh^T H^{-1} \hh}{2}\right) 
			\frac{d^2 \hh} {2\pi|H|^{1/2}}.
\end{equation}
Before we proceed further, the second derivative tensor can be described more conveniently
with the two eigenvalues $\lambda_1,\lambda_2$ and a rotation angle,
$\phi$, as follows:
\begin{eqnarray}
	g_{11} =& \lambda_1 \cos^2\phi +\lambda_2 \sin^2\phi,\\
	g_{12} =& (\lambda_1-\lambda_2) \sin\phi \cos\phi,\\
	g_{22} =& \lambda_1 \sin^2\phi + \lambda_2 \cos^2\phi.
\end{eqnarray}
For simplicity let us introduce the dimensionless variables 
$x=(\lambda_1+\lambda_2)/\sigma_2$, the trace of the second derivative tensor, 
$y=(\lambda_1-\lambda_2)/\sigma_2$, and $z = f/\sigma_0$. The Jacobian of the 
transformation from $(f,g_{11},g_{12},g_{22})$ to $(z,x,y,\phi)$ is
\begin{equation}
		J = \sigma_0\sigma_2^3 y/2.  
\end{equation} 
Let us also introduce the dimensionless $\gamma$ and the characteristic scale 
$R$ as
\begin{equation}
	\gamma = \frac{\sigma_1^2}{\sigma_0\sigma_2},
		\qquad R^2 = \frac{\sigma_1^2}{\sigma_2^2}.
\end{equation}
The quadratic form containing $\vv$ in the exponent can be written with the new 
variables as
\begin{equation}
	Q = v^T C^{-1} v = \left(
		2 y^2	+\frac{x^2 + 2\gamma x z + z^2}{1-\gamma^2}\right).
\end{equation}
The determinants of $C$ and $H$ are 
\begin{equation}
	|C| = \frac{1}{64}(1-\gamma^2)\sigma_0^2\sigma_2^6,\qquad
	|H| = \frac{1}{4} \sigma_1^2.
\end{equation}
In these variables, the unconstrained probability distribution for 
$(x,y,z,\phi)$ becomes
\begin{equation}
	dP = \frac{4y}{(2\pi)^2 \sqrt{1-\gamma^2}} \exp(-\frac{1}{2}Q) 
		\,dx\ dy\ dz\ d\phi.
\end{equation}
In order to properly handle the symmetries of the problem, we can assume that 
$\lambda_1\geq\lambda_2$. Then still any $(\lambda_1,\lambda_2)$ pair can be 
mapped onto itself by a 180 degree rotation, so the valid range of $\phi$ is 
$(0,\pi)$.  Since none of the terms depend on $\phi$, we can integrate 
over $\phi$, resulting in
\begin{equation}
	dP =  
		 \exp\left[-\frac{x^2 + 2\gamma x z + z^2}{2(1-\gamma^2)}\right] 
			\frac{\,dx\ dz}{2\pi \sqrt{1-\gamma^2}}
		\left( e^{-y^2}\,2y\ dy\right).
\end{equation}
The constraint $\lambda_1>\lambda_2$ maps onto $0<y$. If we perform the 
integration over $y>0$, and $x,z$ over $(-\infty,\infty)$, we get 1, as we 
should, for the unconstrained probability for a general point.

As we introduce the peak constraints, we need to first consider the impact
on the gradient. The constrained probability distribution
is
\begin{equation}
	dw_x = \exp\left(-\frac{\hh^T H^{-1} \hh}{2}\right) 
			|\det g| \delta^{(2)}[\hh]			
			\frac{d^2 \hh} {2\pi|H|^{1/2}}.
\end{equation}
After integrating over $d^2\hh$ we get the extremum weight
\begin{equation}
	w_x = \frac{|\det g|}{2\pi|H|^{1/2}} =\frac{x^2-y^2}{4 \pi R^2}.
\end{equation}
This will multiply the unconstrained probability for the density of extremal 
points of the random field, 
\begin{equation}
		dn_{\rm{}pk}=\left[\frac{(x^2-y^2)y}{2\pi R^2} e^{-y^2}\,dy\right]
		\exp\left[-\frac{x^2 + 2\gamma x z + z^2}{2(1-\gamma^2)}\right] 
			\frac{\,dx\ dz}{2\pi \sqrt{1-\gamma^2}}.
\end{equation}
For a peak both eigenvalues of the second derivative tensor must be negative. 
In the rotated coordinates $(x,y)$, this means that $x<0$, and $0<y<-x$.
We can easily integrate over the allowed range of $y$ next, yielding
\begin{equation}
	\int_0^{-x} dy\, y\/ (x^2-y^2)\, e^{-y^2} 
		=\frac{1}{2}(x^2-1+e^{-x^2}).
\end{equation}
We are left with
\begin{equation}
	dn_{\rm{}pk} = \frac{(-1 + x^2 +e^{-x^2})}{4\pi R^2}  
	\exp\left[-\frac{x^2 + 2 x z \gamma + z^2 }{2 (1-\gamma^2)}\right]
	\frac{\,dx\ dz}{2\pi \sqrt{1-\gamma^2}}.
\end{equation}
Let us introduce the function $B(x)$ as
\begin{equation}
	B(s,b) = \sqrt{\frac{\pi}{b}} \exp\left({\frac{s^2}{2b}}\right)
	\left[1+ {\rm erf}\left( \frac{s}{\sqrt{2b}}\right)\right].
\end{equation}
Evaluating the integral over $-\infty<x\leq 0$ in Mathematica, we obtain
\begin{equation}
 n_{\rm{}pk}(s,\gamma)= 
		%\exp\left(-\frac{s^2}{2\gamma^2}\right) 
		\frac{e^{-\frac{s^2}{2\gamma^2}}}{8\pi^2R^2}
		\Bigg[
			(1-\gamma^2)s +[s^2 - \gamma^2 (1 + s^2)] B(s,1)
			+B(s, 3- 2\gamma^2),
		\Bigg]
\end{equation}
with
\begin{equation}
	s = \frac{\gamma z}{\sqrt{1-\gamma^2}}.
\end{equation}

We need to evaluate the shape parameter $\gamma$. Assume that the window 
function applied to the random field is a Gaussian with a scale $a$,
\begin{equation}
	w(r) = \frac{1}{2\pi a^2} \exp\left(-\frac{r^2}{2 a^2}\right).
\end{equation}
Its Fourier transform is also a Gaussian,
\begin{equation}
	W(k) = \exp\left(-\frac{k^2a^2}{2}\right).
\end{equation}
We model the sky noise as a white noise with a flat spectrum. Thus the
correlations in the measured random field are determined by the window function,
i.e.,
\begin{equation}
	P(k) = A\,|W(k)|^2 = A \exp\left(-k^2a^2\right).
\end{equation}
With this power spectrum it is straightforward to compute the scale and the shape 
parameters as
\begin{equation}
	\gamma^2 = \frac{1}{2}, \qquad R^2 = \frac{a^2}{2}.
\end{equation}
